Schema Change: Challenge Problems

### 2024-05-13 paper.tex
rewrite divergence control explanation

\subsection{Context}
Schema change is usually concerned with migrating old data into a new schema. But sometimes we also want to migrate in the reverse direction: new data into an old schema. \citet{Cambria} discuss various ways that code changes can require bidirectional migration. Instead in this problem we focus on changes made by end-users. Users frequently make copies of structured documents like spreadsheets and then diverge them by altering both data content and schema. In the case of spreadsheets schema changes include rearrangements to the structure of rows and columns as well as changes to formulas. The problem is that the users want to transfer such data and schema changes between these divergent copies, and they want to pick and choose which of these changes to transfer. In practice such transfer is done manually through copy \& paste. \citet{Basman19} has documented an ecology of emailed spreadsheets. \citet{Burnett14} distilled field observations of these practices in the story of Frieda, which we quote here:

>>>>

\subsection{Context}
Schema evolution becomes more challenging when data, code, and schema cannot all be updated together in an atomic way. In these situations, different versions must coexist and interoperate with one another: for example, edits made in one schema must be applied in another, or code written against one schema must run against data stored in another.

Such situations are common in software engineering. Web backend architectures often put data and code on separate machines, which makes it impossible to update both atomically. To perform upgrades, teams must manually perform multi-step [zero-downtime deployment processes](https://planetscale.com/blog/zero-downtime-rails-migrations-planetscale-rails-gem) to step the database and the code servers forward gradually such that they are compatible with one another at any given step. One example would be deploying a change to add a database column before deploying the code that uses the column.

In the simplest cases, divergence can last only for a brief moments in between zero-downtime deployment steps. But divergence can also occur across longer timescales, especially in situations where the different parts of a system are not all centrally controlled. For example, a public web API may need to support applications using old versions of the API for months or years, since it is not realistic to force all consumers to upgrade immediately. Divergence can even be indefinite—\citet{Cambria} addresses collaborative applications where different users may want to collaborate on shared data through different tools using different schemas, with no intention of ever converging.

Divergence does not just occur in software engineering; it also appears in end-user workflows. Users frequently make copies of structured documents like spreadsheets and then diverge them by altering both data content and schema. In the case of spreadsheets schema changes include rearrangements to the structure of rows and columns as well as changes to formulas. The problem is that the users want to transfer such data and schema changes between these divergent copies, and they want to pick and choose which of these changes to transfer. In practice such transfer is done manually through copy \& paste. \citet{Basman19} has documented an ecology of emailed spreadsheets. \citet{Burnett14} distilled field observations of these practices in the story of Frieda, which we quote here:
